---
title: "PA1_template"
author: "Dave Cartier"
date: "1/10/2022"
output: html_document
---
# Reproducible Research: Peer Assignment 1

```{r, echo=TRUE}
## Unzip the activity zip file and load the data

datafile <- "activity.csv"
if (!file.exists(datafile)) {
	unzip(zipfile = "activity.zip")
}

activity <- read.csv(datafile, header = TRUE)

```

## Load all libraries that will be used in this analysis

```{r, echo=TRUE}
suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))

```

## What is mean total number of steps taken per day?

```{r, echo=TRUE}
## Calculate the total number of steps taken per day

Total_Steps <- activity %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(TotalStep = sum(steps, na.rm=TRUE))

## Show the total number of step taken per day as a histogram chart

g <- ggplot(data=Total_Steps, aes(TotalStep))
g + geom_histogram() + labs(title = "Total Number of Step Taken per Day")

## Calculate mean and median total number of steps

Mean <- mean(Total_Steps$TotalStep)
Median <- median(Total_Steps$TotalStep)

```

The mean number of steps per day is `r format(round(as.numeric(Mean), 0), nsmall = 1, big.mark = ",")`.
The median number of steps per day is `r format(round(as.numeric(Median), 0), nsmall = 1, big.mark = ",")`.

## What is the average daily activity pattern?

```{r, echo=TRUE}
## Create a time series plot of the 5 minute intervals and the average number of steps taken

Interval_Steps <- activity %>%
  dplyr::group_by(interval) %>%
  dplyr::summarise(AverageStep = mean(steps, na.rm=TRUE))

a <- ggplot(data=Interval_Steps, aes(x=interval, y=AverageStep))
a + geom_line()

## Identify which 5 minute interval has the highest number of average steps

MaxStep <- max(Interval_Steps$AverageStep)

#MaxInterval <- Interval_Steps %>%
#  dplyr::filter(AverageStep == MaxStep) %>%
#  dplyr::select(interval)

MaxInterval <- Interval_Steps[
  Interval_Steps$AverageStep == MaxStep,
]$interval

```

The 5-minute interval that contains the highest daily average of steps is `r MaxInterval`.

## Imputing missing values

```{r, echo=TRUE}

##  Calculate the number of missing values (coded as NA)

Miss <- sum(is.na(activity$steps))

## Strategy is to replace missing values with the mean for that 5 minute interval (calculated above)

## Create a new dataset replacing NA activity values with the mean 5 minute interval

New_Activity <- activity
  for(r in 1:nrow(New_Activity)) {
    if(is.na(New_Activity$steps[r])){
      i <- New_Activity$interval[r];
      avg <- Interval_Steps[Interval_Steps$interval==i,]$AverageStep
      New_Activity$steps[r] <- avg
    }
  }

## Calculate the total number of steps taken per day

New_Total_Steps <- New_Activity %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(TotalStep = sum(steps, na.rm=TRUE))

## Show the total number of step taken per day as a histogram chart

New_g <- ggplot(data=New_Total_Steps, aes(TotalStep))
New_g + geom_histogram() + labs(title = "New Total Number of Step Taken per Day")

## Calculate mean and median total number of steps

New_Mean <- mean(New_Total_Steps$TotalStep)
New_Median <- median(New_Total_Steps$TotalStep)

```

The number of missing values (NA) in the activity dataset is `r Miss`.

The new mean number of steps per day is `r format(round(as.numeric(New_Mean), 0), nsmall = 1, big.mark = ",")` compare to the original mean `r format(round(as.numeric(Mean), 0), nsmall = 1, big.mark = ",")`.

The new median number of steps per day is `r format(round(as.numeric(New_Median), 0), nsmall = 1, big.mark = ",")` compare to the original mean `r format(round(as.numeric(Median), 0), nsmall = 1, big.mark = ",")`.

Replacing the NA values with the average steps at an interval level results in both a higher mean and median of step per day.

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}

## Create a weekday indicator

New_Activity <- New_Activity %>%
  dplyr::mutate(weekday = weekdays(as.Date(date))) %>%
  dplyr::mutate(day_type = case_when((weekday == 'Saturday') ~ 'Weekend',
                                     (weekday == 'Sunday') ~ 'Weekend',
                                     TRUE ~ 'Weekday'))

## Create a panel plot comparing weekday average and weekend average

W_Ave_Steps <- New_Activity %>%
  dplyr::group_by(day_type, interval) %>%
  dplyr::summarise(AveStep = mean(steps, na.rm=TRUE))

W_g <- ggplot(data=W_Ave_Steps, aes(interval, AveStep))
W_g + geom_line(color = 'red', size = 1) + labs(title = "New Total Number of Step Taken per Day Type") + facet_wrap(~ day_type)
  

```

